Racial and Ethnic Differences in ETI

### 2023-09-22 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
KM and first cox model done

## Overall by year

```{r}
annual %>%
  select(eti_elig, ReviewYear) %>%
  tbl_summary(
    by = ReviewYear, missing_text = "Missing",
    type = all_dichotomous() ~ "categorical"
  ) %>%
  bold_labels()
```

## By race and year

```{r results='asis'}
invisible(lapply(names(years), function(n) {
  cat("\n")
  cat("###", n)
  cat("\n")
  d <- years[[n]]
  t <- d %>%
    select(eti_elig, race) %>%
    tbl_summary(
      by = race, missing_text = "Missing",
      type = all_dichotomous() ~ "categorical"
    ) %>%
    bold_labels()
  print(t)
  cat("\n")
}))
```

@@

```{r}
#| cache: true
# Most basic model of lung function over time by race
# Try a few different models
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
label(means) = list("Race/Ethnicity","Year","Mean ppFEV","SE","df","Lower CL", "Upper CL")
gt(means) %>% fmt_number(emmean:last_col(),decimals = 3)
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID),alpha=0.005,linewidth=0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean,color=race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean,group = race,color=race)) +
  geom_errorbar(mapping=aes(x = ReviewYear,ymin = emmean - SE, ymax = emmean + SE,color=race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID),alpha=0.005,linewidth=0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean,color=race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean,group = race,color=race)) +
  geom_errorbar(mapping=aes(x = ReviewYear,ymin = emmean - SE, ymax = emmean + SE,color=race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity") +
  facet_wrap(~race)
```

@@

```{r}
#| cache: true
t = annual %>% select(eDWID,ReviewYear,mean_ppFEV) %>% na.omit(.)
model = list(mean_ppFEV ~ 0 + ReviewYear, 
              ~ 0 + ReviewYear)
fit = mcp(model,data = t,cores=4,chains = 4)
plot(fit)
```

>>>>

The following tables have been aggregated by year, and therefore may miss some information regarding eligibility and ETI start. For example, if someone turned 12 on 12/30/2019, they would technically be considered eligible for ETI, even though realistically we probably wouldn't expect them to start that year.

## Proportions

### Overall by year

```{r}
annual %>%
  select(eti_elig, contains("Vx"), ReviewYear) %>%
  tbl_summary(
    by = ReviewYear, missing_text = "Missing",
    type = all_dichotomous() ~ "categorical"
  ) %>%
  bold_labels()
```

### By race and year

```{r results='asis'}
invisible(lapply(names(years), function(n) {
  cat("\n")
  cat("###", n)
  cat("\n")
  d <- years[[n]]
  t <- d %>%
    select(eti_elig, contains("Vx"), race) %>%
    tbl_summary(
      by = race, missing_text = "Missing",
      type = all_dichotomous() ~ "categorical"
    ) %>%
    bold_labels()
  print(t)
  cat("\n")
}))
```

## Survival analysis

For the following analysis, the start time ($t=0$) was the approximate date that each person became eligible for ETI. Because we don't have exact dates of birth, I calculated an approximate DOB based on age and encounter dates.

```{r}
# Create survival dataset
surv_dat <- encounter %>%
  mutate(stop = as.numeric(encounterdate - eti_elig_date)) %>%
  filter(stop >= 0) %>%
  group_by(eDWID) %>%
  mutate(start = lag(stop)) %>%
  select(eDWID, race, encounterdate, encounterage, eti_elig_date, Vx445comb, start, stop) %>%
  filter(stop <= first(stop[Vx445comb == "Yes"]))
surv_dat$Vx445comb <- as.numeric(factor(surv_dat$Vx445comb, levels = c("No", "Yes")))
# Filter for KM
km_dat <- surv_dat %>%
  group_by(eDWID) %>%
  slice_max(stop)
kmfit <- survfit(Surv(stop, Vx445comb) ~ race, data = km_dat[km_dat$race != "Asian", ])
autoplot(kmfit) + theme_bw()
# Cox PH model
fit <- coxph(Surv(start, stop, Vx445comb) ~ race, id = eDWID, data = surv_dat)
```

@@

```{r}
#| cache: true
# Most basic model of lung function over time by race
# Try a few different models
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
label(means) <- list("Race/Ethnicity", "Year", "Mean ppFEV", "SE", "df", "Lower CL", "Upper CL")
gt(means) %>% fmt_number(emmean:last_col(), decimals = 3)
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID), alpha = 0.005, linewidth = 0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID), alpha = 0.005, linewidth = 0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity") +
  facet_wrap(~race)
```

@@

```{r}
#| cache: true
#| eval: false
t <- annual %>%
  select(eDWID, ReviewYear, mean_ppFEV) %>%
  na.omit(.)
model <- list(
  mean_ppFEV ~ 0 + ReviewYear,
  ~ 0 + ReviewYear
)
fit <- mcp(model, data = t, cores = 4, chains = 4)
plot(fit)
```
